Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation

### 2013-04-03 manuscript/manuscript.tex
update figures and tables

    \Procedure{Call SVs}{}
    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
    \EndFunction
    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
    \EndFunction
    \EndProcedure
  \end{algorithmic}
\label{cloudbreakAlgorithm}
\caption{The algorithmic framwork for SV calling in MapReduce.}
\end{algorithm}

@@

We have implemented an indel detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak. Our implementation defines the three application-specific functions described above as:
\begin{description}
\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus. Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

@@

@@

\begin{figure}[b]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
\label{chr2DeletionsRoc}
\end{figure}

@@

Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the perfomance characteristics of algorithms at varying levels. First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required. Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel. Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches. Pindel exlusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  &  47 (  7)  &  52 (\textbf{  3}) &  \textbf{ 57} (\textbf{  5}) & \textbf{ 12} (\textbf{  4}) & \textbf{ 15} (\textbf{0}) \\ 
  Breakdancer &  52 ( 10)  &  49 (\textbf{  3}) &   49 (0) &   7 (0) &  14 (\textbf{0}) \\ 
  GASVPro     &  35 (  4)  &  26 (0) &   26 (0) &   2 (0) &   6 (\textbf{0}) \\ 
  DELLY       &  22 (  2)  & \textbf{ 56} (  2) &   40 (0) &   8 (0) &  12 (\textbf{0}) \\ 
  Pindel      & \textbf{ 60} (\textbf{ 34})  &  16 (0) &   41 (  1) &   1 (0) &  12 (\textbf{0})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsFDR10}
\end{table}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak    &  \textbf{0.532} & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
  DELLY           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (  1) &   9 (0) &  17 (  1) \\ 
  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsMaxSensitivity}
\end{table}

\begin{figure}[b]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
\label{chr2InsertionsRoc}
\end{figure}

An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}. The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}. Cloudbreak shows excellent precision, although recall is low for all four aproaches. Cloudbreak again identifies the most small variants. Pindel is the only tool which can identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
  \hline
  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses). The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
\label{chr2InsertionPreds}
\end{table}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
\label{NA18507DeletionsRoc}
\end{figure}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
  \hline
Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 19}) \\ 
Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
  DELLY & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  3) \\ 
  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found. The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
\label{NA18507DeletionPreds}
\end{table}

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
\label{NA18507InsertionsRoc}
\end{figure}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
Total Number & & & 536 & 114 & 45 & 1 \\
  \hline
Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
\label{NA18507InsertionPreds}
\end{table}

@@

We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance. We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements in both the simulated and NA18507 data sets. Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above. In the NA18507 data set, Cloudbreak identifies 16\% more repeat-overlapping variants than the nearest competitor, Breakdancer. Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible. Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 553 & 7851 \\ 
  \hline
  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{deletionRepmaskpreds}
\end{table}

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 133 & 270 & 341 & 355 \\ 
  \hline
  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
\label{insertionRepmaskpreds}
\end{table}

@@

We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots. Not including alignment time, we were able to process the Chromosome 2 simulated data in under 10 minutes, and the the NA18507 data set in under an hour. For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300. The bulk of Cloudbreak's execution is spent in the feature generation step. Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.

>>>>

    \Procedure{Call SVs}{}
    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
    \EndFunction
    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
    \EndFunction
    \EndProcedure
  \end{algorithmic}
\label{cloudbreakAlgorithm}
\caption{The algorithmic framework for SV calling in MapReduce.}
\end{algorithm}

@@

We have implemented an indel detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak. Our implementation defines the three application-specific functions described above as:
\begin{description}
\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure \ref{insert_size_mixes}). Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{/users/cwhelan/Documents/gene_rearrange/insert_size_mixtures.pdf}
\caption{Illustration of insert size mixtures at individual genomic locations. A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings. B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes. C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected. D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
\label{insert_size_mixes}
\end{figure}

@@

A diagram of the Cloudbreak algorithm working on a simple example is shown in Figure \ref{algorithm_example}.

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
\caption{An example of the Cloudbreak MapReduce algorithm. A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads. Reducers gather all alignments for both reads in each pair. B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair. Only one genomic location is diagrammed here for simplicity. Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes. C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
\label{algorithm_example}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
\label{chr2DeletionsRoc}
\end{figure}

@@

Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels. First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required. Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel. Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches. Pindel exclusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.

\begin{table}
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  &  47 (  7)  &  52 (\textbf{  3}) &  \textbf{ 57} (\textbf{  5}) & \textbf{ 12} (\textbf{  4}) & \textbf{ 15} (\textbf{0}) \\ 
  Breakdancer &  52 ( 10)  &  49 (\textbf{  3}) &   49 (0) &   7 (0) &  14 (\textbf{0}) \\ 
  GASVPro     &  35 (  4)  &  26 (0) &   26 (0) &   2 (0) &   6 (\textbf{0}) \\ 
  DELLY       &  22 (  2)  & \textbf{ 56} (  2) &   40 (0) &   8 (0) &  12 (\textbf{0}) \\ 
  Pindel      & \textbf{ 60} (\textbf{ 34})  &  16 (0) &   41 (  1) &   1 (0) &  12 (\textbf{0})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsFDR10}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak    &  \textbf{0.532} & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
  DELLY           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (  1) &   9 (0) &  17 (  1) \\ 
  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsMaxSensitivity}
\end{table}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
\label{chr2InsertionsRoc}
\end{figure}

An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}. The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}. Cloudbreak shows excellent precision, although recall is low for all four approaches. Cloudbreak again identifies the most small variants. Pindel is the only tool which can identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.

\begin{table}
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
  \hline
  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses). The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
\label{chr2InsertionPreds}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
\label{NA18507DeletionsRoc}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
  \hline
Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 19}) \\ 
Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
  DELLY & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  3) \\ 
  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found. The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
\label{NA18507DeletionPreds}
\end{table}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
\label{NA18507InsertionsRoc}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
Total Number & & & 536 & 114 & 45 & 1 \\
  \hline
Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
\label{NA18507InsertionPreds}
\end{table}

@@

We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance. We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements in both the simulated and NA18507 data sets. Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above. In the NA18507 data set, Cloudbreak identifies 15\% more repeat-overlapping variants than the nearest competitor, Breakdancer. Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible. Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.

\begin{table}
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 553 & 7851 \\ 
  \hline
  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{deletionRepmaskpreds}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 133 & 270 & 341 & 355 \\ 
  \hline
  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
\label{insertionRepmaskpreds}
\end{table}

@@

We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots. Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes. For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300. The bulk of Cloudbreak's execution is spent in the feature generation step. Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
